# **Chapter6:Analysis on Longitudinal data**

**1. Implementing the analyses of Chapter 8 of MABS using the RATS data.**

**Introduction**

- The RATS data set is from  a nutrition study conducted in three groups of rats. The groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately) weekly, except in week seven when two recordings were taken) over a 9-week period. The question of most interest is whether the growth profiles of the three groups differ.

- First, I will read the RATS dataset into R and have a look at it.
```{r}
#  Reading the RATS data set into R. 
RATSL <- read.table("data/RATS.csv",sep  =",",header = TRUE)

str(RATSL)
summary(RATSL)
names(RATSL)
head(RATSL)

# Factor ID & Group in RATS dataset
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

```

**Graphical Displays of Longitudinal Data**

Graphical displays of data are almost always useful for exposing patterns in the data. Following Kimmo's book, I will next plot the data with Time on x axis ant the variable rats on the y axis. 

The variable "rats" gives the rat's body weight (in grams), When I wrangled the data I wasn't quite sure yet what the variable stood for so that explains the variable name that is not so great.
```{r}
#Access the package ggplot2
library(ggplot2)
# Draw the plot
ggplot(RATSL, aes(x = Time, y = rats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$rats), max(RATSL$rats)))

#An important effect we want to take notice is how the rats who have higher value of "rats" at the beginning tend to have higher values throughout the study. This phenomenon is generally referred to as tracking.The tracking phenomenon can be seen more clearly in a plot of the standardized values of each observation.

# Standardise the variable rats
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdrats = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

glimpse(RATSL)

# Plot again with the standardised bprs
ggplot(RATSL, aes(x = Time, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized rats")
```
We can see that there are big differences in the weight of the rats (value of "rats")  between the 3 groups, this is mostly explained by the fact, that the rats in different groups had were of different weight from the beginning. 

The rats in group 1 were lightest, under 300 grams, most of the rats in group 2 under 500 grams (except for the one that is much bigger than the others) and most of the rats in group 3 over 500 grams.

It is hard to compare the effects of the nutrition on the rats since the groups were so different from the beginning. Still it seems that the weight of the rats has gone up during the study in all the groups.



**Summary Measure Analysis of Longitudinal Data**

With large numbers of observations, graphical displays of individual response profiles are of little use and investigators then commonly produce graphs showing average (mean) profiles for each treatment group along with some indication of the variation of the observations at each time point, in this case the standard error of mean.

```{r}
library(dplyr)
library(tidyr) 

# Number of Time
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(rats), se = sd(rats)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(rats) +/- se(rats)")


```

In the case of the RATS dataset, I do not think that this summary graph offers much more information compared to the previous graphs. We can see that the standard error of group 2 is bigger than in the other groups, probably explained to some part by the one value that differed from the others,


**2.Implementing the analyses of Chapter 9 of MABS using the BPRS data.**

- First, I will read the BPRS dataset into R and have a look at it.

In the BPRS data 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.

```{r}
#  Reading the RATS data set into R. 
BPRSL <- read.table("data/BPRS.csv",sep  =",",header = TRUE)

str(BPRSL)
summary(BPRSL)
names(BPRSL)
head(BPRSL)

# Factor treatment & subject in BPRS data set
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

```

**Linear Mixed Effects Models for Repeated Measures Data**
random intercept model
random intercept and slope model

To see how linear mixes effects differ, I will use the  BPRS data. The question of most interest is, if the profiles of bprs values differ between the treatment groups.

I will first ignore the repeated-measures structure of the data and assume that all observations are independent of each other.

I have a data of 360 observations on 5 different variables that can easily be analyzed by using multile linear regression.

First, I will plot the data identifying the observations in each group, but ignoring the longitudinal nature of the data.

```{r}

library(ggplot2)
# Plot the BPRSL data
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) + geom_line()

```
My graph is not the easiest to interpret (not sure why as I tried to do it as instructed!), but there seems to be quite a bit variation in the value of bprs.

```{r}
#ggplot(BPRSL, aes(x = week, y = bprs, group = #treatment)) + geom_line()

```
From the second graph it is easier to see that the two treatment groups differ from eachother in the value of bprs most whitin the first week, but not so much later on.

**Fitting Linear Mixed Models to the BPRS Data**

Continuing to ignore the repeated-measures structure of the data, I will fit a multiple linear regression model  to RATSL Data with bprs as Response Variable, and treatment and week as Explanatory Variables

```{r}
library(dplyr)
library(tidyr)

# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)



```
Treatment group 1 has a strong positive effect for the value of bprs and treatment group2 has a smaller positive effect. The result is statistically signifigant for treatment group 1.

week has a negative correlation with bprs. The results are statistically signifigant.

**The random Intercept Model**

The previous model assumed independence of the repeated measures of bprs, and this assumption is highly unlikely. So, now I will move on to consider both some more appropriate graphics and appropriate models.

I will first fit the random intercept model for the same two explanatory variables:week& treatment. Fitting a random intercept model allows the linear regression fit for each rat to differ in intercept from other rats.

```{r}

# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```
The standard error for week gets smaller, which reflects the point that assuming independence will lead to the standars error of a whit-in subject covariate such as week being larger than it should be, because  of ignoring the likely within subject dependences, which will reduce the error variance in the model.

The standars error of treatment gets larger. These are between subject-effects and the reason for the smaller
standard errors with the independence model is that the effective sample size for estimating these effects is less than the actual sample size because of the correlated nature of the data, and so the estimates for the independencemodel are unrealistically precise.

**Random Intercept and Random Slope Model**

Fitting a random intercept and random slope model allows the linear regression fits for each individual to differ in intercept but also in slope. This way it is possible to account for the individual differences in the bprs profiles of subjects, but also the effect of time.
```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

#Finally, we can fit a random intercept and slope model that allows for a group × time interaction.

# create a random intercept and random slope model
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of RATSL
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Tweek", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to BPRSL
RATSL <- BPRSL %>%
  mutate(Fitted)

# draw the plot of RATSL
ggplot(RATSL, aes(x = Time, y = Fitted, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")

```

