# **Chapter5:Dimensionality reduction techniques**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**1.**  Showing a graphical overview of the data and show summaries of the variables in the data.
```{r}

human_ <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", sep=",", header=TRUE)

summary(human_)


library(GGally)
library(dplyr)
library (corrplot)
# visualize the 'human_' variables
ggpairs(human_)

# compute the correlation matrix and visualize it with corrplot
cor(human_) %>% corrplot


```

There are large differences in the variables between different countries. Here are some observations: 

- life expectancy varies between 49 and 83,5 years
- maternal mortality rate varies between 1 and 1100
- female labour participation compared to men varies between 0.1857 and 1.0380
- female secondary education compared to men varies between 0.1717 and 1.4967
- GNI varies between 581 and 123124

There is a **positive correlation** between expected education and life expectancy
also a **positive correlation** between adolescent birthrate and maternal mortality

There is a **negative correlation** between maternal mortality and life expectancy
There is a **negative correlation** between maternal mortality and expected education
There is also a **negative correlation** between adolescent birthrate and life expectancy
There is also a **negative correlation** between adolescent birthrate and expected education

We can note that higher female secondary education is positively correlatad with things like educational expectation, life expectacy and GNI and negatively correlated with adolescent birthrate and maternal mortality.

**2.** Next, I will perform **principal component analysis (PCA) on the not standardized human data**. I will show the variability captured by the principal components and draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables. 

```{r}

# print out summaries of the human_ variables
summary(human_)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2)

```


GNI has very high variance and captures most of the variability in the data

**3.** Next, I will standardize the variables in the human_ data and repeat the above analysis. 

The standardization will set the mean of the variables to zero thus creating a different biplot.
```{r}
# standardize the variables
human_std <- scale(human_)

# print out summaries of the standardized variables
summary(human_std)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2)
```


**4.** Based on the biplot drawn after PCA on the standardized human data we can note the following: 

- Female labour force participation and female parliament representation are pointing to the same direction as the second principal component so they contributre to that dimension. 

- All the other features are pointing to the direction of the first principal component, thus contributing to that dimension.

**5.** I will load the tea dataset from the package FactoMineR and explore the data briefly. 

The tea dataset contains 36 variables and 300 observations

```{r}
# access the Factominer, ggplot2, dplyr and tidyr packages
library(FactoMineR)
library(ggplot2)
library(dplyr)
library(tidyr)

# load the tea data and exploring the data
data("tea")
str(tea)
dim(tea)
summary(tea)

# visualize the dataset
boxplot(tea[,0:18], main='Multiple Box plots')
boxplot(tea$age)
boxplot (tea[,20:36], main='Multiple Box plots')

```


Next, I will create a new data set "tea_time" and do **Multiple Correspondence Analysis (MCA)** on it and comment on the outputs.

```{r}

# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- dplyr::select(tea, one_of(keep_columns))

# look at the summaries and structure of the data
summary(tea_time)
str(tea_time)

# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))


# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"))
plot(mca, invisible=c("var"))


#adjusting the code by adding argument habillage = "quali". This will differentiate the different variables with different colors, making the visualization easier to interpret.
plot(mca, invisible=c("ind"), habillage = "quali")


# multiple correspondence analysis also visualizes the analysis by default
mca <- MCA(tea_time)

```
